Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-15 aistats.tex
Merge Hanzely and Dragomir

\subsection{Asymptotic Rate}
\label{ssec:asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
We now detail a classical asymptotic result.
Proofs are in App.~\ref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.

Statistics typically give results on $\nat$, but the MAP~\eqref{eq:defMAP} is more simply expressed with $\meanp$, so let us focus on $\bregmanconj$.
Bregman divergences are locally quadratic, as seen via a second order Taylor expansion
\alignn{
    \textstyle \bregmanconj(\m^* ; \m)
    &\textstyle = \frac{1}{2}\norm{\m^* - \m}^2_{\mF}
    + O(\norm{\m - \m^*}^3),
    \label{eq:bregmanTaylor}
}
where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the Hessian of the entropy at the optimum. It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
\aligns{
    \mF
    :=\nabla^2\conj(\m^*)
    = \nabla^2\logpart(\nat^*)^{-1}
    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
}
Plugging the MLE~\eqref{eq:defMLE} or MAP~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, we have
\begin{align}
	\label{eq:asymptote}
	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right )
	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
\end{align}
Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible for large $n$.
For well-specified models, this asymptote is also independent of the optimum $\meanp^*$ or $\mF$.
Next, we give another example for which we get a rate matching~\eqref{eq:asymptote}.

\subsection{Quadratic Case}
\label{ssec:quadratic}
As another classical reference point, we consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
For instance, this is the log-partition of a Gaussian with known variance $I$,
\[
	\cX=\real^d,\quad \nu(dx) = \exp\paren{\textstyle \half[-\|x\|^2]} dx,\quad T(x)=x.
\]
In this case, $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances since
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples, we can break down the MLE into individual point's contributions:
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}
	=\frac{\Var(T)}{2n}
	=\frac{d}{2n}.
\end{align}
Adding a reference mean $\m_0$ to get the MAP yields
\begin{align}
		\!\!\!\!\! \expect{\half \norm{\m^* -   \MAPm^\text{MAP}}_2^2} \!
	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.\!
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
	\label{eq:MAP_quadratic}
\end{align}
We see here a variance term defining the $\frac{d}{2n}$ asymptote and a bias term in $O(n^{-2})$. However, this result does not generalize well to other families, unless we make restrictive assumptions on $\conj$. % we can relate $\bregmanconj$ to a norm or a quadratic.
% Can we generalize these results to other families?

@@

\subsection{Locally Quadratic Case}
\label{ssec:local-quadratic}
From Taylor expansion~\eqref{eq:bregmanTaylor},
we know that all Bregman divergences are locally quadratic.
Under some assumptions, such as self-concordance\footnote{
In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
} of $\conj$ \citep[Ch.~4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in App.~\ref{app:self-concordant}.
\begin{proposition}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$.}
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21
	\implies
	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2.
}
\end{proposition}
To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
For the MLE, the proof of~\eqref{eq:asymptote} from~\eqref{eq:bregmanTaylor} yields $\expecti{\norm{\m^*-\hat \m_n}^2_{\mF}} = \frac{d}{n}$ in general, so a sufficient condition is $n \geq 25 d$.
For MAP, transforming~\eqref{eq:MAP_quadratic}, we get the sufficient condition \TODO{FIX THIS!} $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
This means that on average, we need $25$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.

The entropy $\conj$ is self-concordant for several common families.
For instance when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.~\ref{app:self-concordant}),
as is the case with:
exponential distributions,
Gaussians with known mean,
Laplace with known mean,
Pareto with known minimum value,
or Weibull with known shape $k$.
The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic} -- e.g. categorical and Dirichlet distributions.

@@

These results partially answer~\eqref{problem}.
Partially, because the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$.
None of them apply to small $n$.
That is one caveat. Another one is the Gaussian example we saw in \S\ref{ssec:gaussian}, which is not self-concordant.
A full solution to~\eqref{problem} would address both of these caveats.
Informed by the properties we have seen so far, let us investigate a general decomposition of the Bregman that could guide us towards a solution.
%If no current approach solves~\eqref{problem}, then let us formulate a conjecture informed by the examples and partial solutions we have s.

@@

Recently, these assumptions have been relaxed to the $\stgcvx$-strong convexity and $\smooth$-smoothness of $f$
\emph{relative} to a reference function $\logpart$, defined as
\aligns{
	\stgcvx \cB_{A}(x, y)
	\leq
	\cB_f(x,y)
	\leq
	\smooth \cB_A(x,y) \; .
}
When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent.
Those conditions ensure the linear convergence of MD with mirror map $\nabla A$,
extending the theory to non-smooth functions $f$ and non-strongly convex potential $\logpart$
\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.

For exponential families, the log-likelihood perfectly fit into this framework, as
\aligns{
	f(\theta) = A(\theta) - \expect{\lin{T(X), \theta}}
}
is both $1$-smooth and $1$-strongly convex relative to $A$.
So we are left with finding an applicable convergence rate for SMD under relative assumptions.
As we investigate in the next section, several pieces of work have addressed this question, under various assumptions on the stochasticity \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.

\subsection{Bounding the Randomness}
\TODO{compress this section}

To analyze stochastic algorithms, one also need to quantify the randomness.
Many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}.
We now review attempts at analyzing SMD under relative smoothness assumption, in the light of  the MAP example.
See \cref{tbl:assumptions} for a summary.

\textbf{Caveat:} these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging.

\begin{table}[t]
\begingroup
%\renewcommand*{\arraystretch}{1.25}%
\newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
\newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
\centering
\caption{Summary of results for SMD
under relative smoothness and relative strong convexity assumptions.
Barrier tells whether the assumption can hold with barrier objectives.
$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
}
\begin{tabular}{lccc}
%{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
\toprule
Boundedness & Barriers &  $\lr_n \sim \inv{n}$ & Last iterate \\
\midrule
%Strongly convex and bounded variance\newline
%\citep[e.g.,][]{??}
%&
%$
%\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
%\quad
%\expect[x]{\norm{\m - x}}^2\leq\sigma^2
%$
%\\
Var. on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
& \redxmark & \greencmark  & \redxmark
\\
Var. at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
& \redxmark & \greencmark  & \greencmark
\\
Opt. gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
& \greencmark & \redxmark & \greencmark
\\
\bottomrule
\end{tabular}
\label{tbl:assumptions}
\endgroup
\end{table}

\subsubsection{Bounded Variance}
\citet{hanzely2018fastest} assume the covariance between negative gradient and next iterate is upper bounded at every time step: $\forall n \geq 1, \forall \hat \nat_{n-1},$
\alignn{
	\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) \leq \lr_n C
	\label{eq:hanzely}
}
for some constant $C$.
For a Gaussian with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$),
this definition recovers the variance of the stochastic gradient
\[
	\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C.
\]
More generally, if $A$ is $\stgcvx$-strongly convex w.r.t. $\norm{\cdot}$,
the LHS of \cref{eq:hanzely} is bounded by
$\expecti{\frac{1}{\stgcvx}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$.
Then a bound on the variance of stochastic gradients is sufficient for \cref{eq:hanzely}
to hold.
%Another perspective on this covariance is as the expectation of the symmetrized Bregman $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ between stochastic and deterministic updates
%\aligns{
%\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) = \E[\cS(\MAPm - \lr g_n  )]
%}
%thus showing positivity.
Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate with $O(1/n)$ step-sizes,
by applying tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
But in general, this assumption rarely holds.
For the MAP, this translates into $\forall n \geq 1,$
\alignn{
\expect[T_n]{\lin{T_n - \meanp^* , \MAPt}} \leq \lr_n C
}
which fails to hold on the gaussian variance example for instance. Indeed, we cannot control the samples $x_i$, therefore there is a non-zero probability they make $\nabla\conj(\MAPt) = \hat\meanp_t$ arbitrarily close to zero. Since the conjugate of the log partition is $\conj=-0.5\log$, this makes $\MAPt$ unbounded.

%  \TODO{ DAMIEN short proof or formalize.}

\subsubsection{Bounded Variance at the Minimum}

\citet{dragomir2021fast} use the weaker assumption
\alignn{
	\expect[\tilde g]{
		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
	} \leq 2 \lr^2 C \; ,
	\quad
	\forall n \geq 1.
	\label{eq:dragomir}
}
In words, this bounds the variance of the gradients at the minimum, but with a metric that depends on the iterate $\MAPm$, and this bound should hold for every iterate we may encounter on the stochastic trajectory.
% As observed by Frederick, this is really weaker than Hanzely, because it is only the first half of their assumption. There is only a factor 2 that's preventing a direct inequality.
Under this assumption, they prove convergence up to a variance ball with a constant step-size.
Their descent lemma \citep[Eq.(12)]{dragomir2021fast}
is the analog of a modern analysis of SGD~\citep[Th.3.2]{gower2019sgd},
meaning we can plug in a $O(1/n)$ decreasing step-size and obtain a $O(1/n)$ convergence rate.

This seems to be a solution to our problem, but it suffers from the same issues as \citet{hanzely2018fastest}.
For barrier objectives, \cref{eq:dragomir} does not hold uniformly on every path of iterates,
because these paths can get arbitrarily close to boundaries where $\bregmanconj$ diverges.

@@

\subsection{Issues with current analyses}

\fdk{They don't hold for problems we care about}

\fdk{Existing techniques only deal with fixed step-sizes and show convergence to a ball}

@@

%\section{Related Work}

%\paragraph{Online Learning}
%Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
%\[
%R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
%\]
%where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$. Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule. Unfortunately however, their rate are worse than the offline MAP or MLE algorithms. For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$. From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$. Also, \citet{dasgupta2007online} prove the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$. This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.

>>>>

\subsection{Asymptote}
\label{ssec:asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
We now detail a classical asymptotic result.
Proofs are in App.\ref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.

Statistics typically give results on $\nat$, but the MAP~\eqref{eq:defMAP} is more simply expressed with $\meanp$, so let us focus on $\bregmanconj$.
Bregman divergences are locally quadratic, as seen via a second order Taylor expansion
\alignn{
    \textstyle \bregmanconj(\m^* ; \m)
    &\textstyle = \frac{1}{2}\norm{\m^* - \m}^2_{\mF}
    + O(\norm{\m - \m^*}^3),
    \label{eq:bregmanTaylor}
}
where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum. It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
\aligns{
    \mF
    :=\nabla^2\conj(\m^*)
    = \nabla^2\logpart(\nat^*)^{-1}
    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
}
Plugging the MLE~\eqref{eq:defMLE} or MAP~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, we have
\begin{align}
	\label{eq:asymptote}
	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right )
	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
\end{align}
Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible for large $n$.
For well-specified models, this asymptote is also independent of the optimum $\meanp^*$ or $\mF$.
Let us see another example for which we get a rate matching~\eqref{eq:asymptote}.

\subsection{Quadratic Case}
\label{ssec:quadratic}
As another classical reference point, we consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
For instance, this is the log-partition of a Gaussian with known variance $I$,
\[
	\cX=\real^d,\quad \nu(dx) = \exp\paren{\textstyle \half[-\|x\|^2]} dx,\quad T(x)=x.
\]
In this case, $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances since
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples, we can break down the MLE into individual point's contributions:
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}
	=\frac{\Var(T)}{2n}
	=\frac{d}{2n}.
\end{align}
Adding a reference mean $\m_0$ to get the MAP yields
\begin{align}
	\expect{\bregmanconj(\meanp^*; \MAPm)}
	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
	\label{eq:MAP_quadratic}
\end{align}
We see here a variance term defining the $\frac{d}{2n}$ asymptote and a bias term in $O(n^{-2})$. However, this result does not generalize well to other families, unless we make restrictive assumptions on $\conj$. % we can relate $\bregmanconj$ to a norm or a quadratic.
% Can we generalize these results to other families?

@@

\subsection{Locally Quadratic Case}
\label{ssec:local-quadratic}
From Taylor expansion~\eqref{eq:bregmanTaylor},
we know that all Bregman divergences are locally quadratic.
Under some assumptions, such as self-concordance\footnote{
In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
} of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in App.\ref{app:self-concordant}.
\begin{proposition}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21
	\implies
	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2.
}
\end{proposition}
To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
This means that on average, we need $5$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.

The entropy $\conj$ is self-concordant for a lot of common families.
For instance when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.\ref{app:self-concordant}),
as is the case with:
exponential distributions,
Gaussians with known mean,
Laplace with known mean,
Pareto with known minimum value,
or Weibull with known shape $k$.
The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic} -- e.g. categorical and Dirichlet distributions.

@@

These results partially answer~\eqref{problem}.
Partially, because the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$.
None of them apply to small $n$.
That is one caveat. Another one is the gaussian example we saw in \S\ref{ssec:gaussian}, which is not self-concordant.
A full solution to~\eqref{problem} would address both of these caveats.
Informed by the properties we have seen so far, let us investigate a general decomposition of the Bregman that could guide us towards a solution.
%If no current approach solves~\eqref{problem}, then let us formulate a conjecture informed by the examples and partial solutions we have s.

@@

Recently, these assumptions have been relaxed to the $\stgcvx$-strong convexity and $\smooth$-smoothness of $f$
\emph{relative} to a reference function $\logpart$, defined as
\aligns{
	\stgcvx \cB_{A}(x, y)
	\leq
	\cB_f(x,y)
	\leq
	\smooth \cB_A(x,y) \; .
}
When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent.
These relative conditions ensure the linear convergence of MD with mirror map $\nabla A$
\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively},
even when $f$ is not smooth and $\logpart$ is not strongly convex.

For exponential families, MAP perfectly fits into this framework, as
\aligns{
	f(\theta) = A(\theta) - \expect{\lin{T(X), \theta}}
}
is both $1$-smooth and $1$-strongly convex relative to $A$.
Our goal is then to find an applicable convergence rate for SMD under relative smoothness.

\subsection{Bounding the Randomness}

To analyze stochastic algorithms, one also need to quantify the randomness.
While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, but they have so far been lacking concrete examples.
We review these analyses in the light of the MAP and provide a summary in \cref{tbl:assumptions}.

\begin{table}[t]
\newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
\newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
\caption{Summary of results for SMD
under relative smoothness and relative strong convexity assumptions.
Each row correspond to one analysis, and each columns answers one question.
($-\log$) does the bound hold for the Gaussian variance example (\S\ref{ssec:gaussian-variance})?
($\lr_n \sim \inv{n}$) does it converge with a $O(\inv{n})$ step-size?
($f$) is the bound in function value, or in reverse Bregman $\bregman(\nat^*,\hat\nat_n)$? 
 ($\hat\nat_n$) is it for the last iterate or an average ?
None of these analysis check all the boxes needed to address~\eqref{problem}.
}
\begin{center}
\begin{tabular}{lcccc}
%{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
\toprule
Boundedness & $-\log$ &  $\lr_n \sim \inv{n}$ & $f$ & $\hat\nat_n$ \\
\midrule
%Strongly convex and bounded variance\newline
%\citep[e.g.,][]{??}
%&
%$
%\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
%\quad
%\expect[x]{\norm{\m - x}}^2\leq\sigma^2
%$
%\\
Var. on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
& \redxmark & \greencmark & \greencmark  & \redxmark
\\
Var. at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
& \redxmark & \greencmark & \redxmark  & \greencmark
\\
Opt. gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
& \greencmark & \redxmark & \redxmark & \greencmark
\\
\bottomrule
\end{tabular}
\end{center}
\label{tbl:assumptions}
\end{table}

%Before we proceed, let $f_X$ be a stochastic estimate of $f = \expecti{f_X}$, with stochastic gradients $g_X(\nat) =\nabla f_X(\nat)$. In our case $f_X(\nat) = - \log p(X\cond \nat)$ and $g_X(\nat) = \m - T(X)$.

\subsubsection{Analogs of the Variance}
let us introduce $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ th symmetrized Bregman induced by $\conj$.
\citet{hanzely2018fastest} assume that the expectation of $\cS$ between stochastic and deterministic updates verifies
\alignn{
\expect[g]{\cS_\conj \paren{
	\hat\m_{n} - \lr g(\hat \nat_{n}), 
	\hat\m_{n}   -\lr \nabla f(\hat \nat_{n})
}} \leq \lr^2 C
\label{eq:hanzely}
}
for all possible iterates $\MAPt$, relevant step-sizes $\lr$ and for some constant $C$.
When $A(\theta) = \frac{1}{2}\norm{\theta}^2$ or when $A$ is $\stgcvx$-strongly convex,
this definition recovers the variance of the stochastic gradient
\[
	\expect[g\!\!]{\inv{\stgcvx}\norm{\nabla f(\theta) - g(\theta)}^2}\leq C.
\]
Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate on function values with $O(1/n)$ step-sizes and tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.

\citet{dragomir2021fast} define the weaker assumption
\alignn{
	\expect[\tilde g]{
		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
	} \leq 2 \lr^2 C \; .
	\label{eq:dragomir}
}
When $\logpart$ is $\stgcvx$-strongly-convex, we recover the variance of the gradients at the optimum
\[
	\expect[g\!\!]{\inv{\stgcvx}\norm{\nabla g(\theta^*)}^2}\leq C.
\]
Using their descent lemma \citep[Eq.(12)]{dragomir2021fast} with the $O(1/n)$ step-size used by \citet[Th.3.2]{gower2019sgd} for SGD, we obtain a $O(1/n)$ convergence rate, on the Bregman with \emph{reversed} arguments $\bregman(\nat^*, \hat\nat_n)$.

These two analyses seem promising for~\eqref{problem}, but none of these assumptions hold in front of barrier objectives such as the $-\log$ from \S\ref{ssec:gaussian-variance}.
Indeed, we cannot control the samples $x_i$, therefore there is a non-zero probability they make $\nabla\conj(\MAPt) = \hat\meanp_t$ arbitrarily close to zero. Since the conjugate of the log partition is $\conj=-0.5\log$, this makes $\MAPt$ unbounded.

@@

@@
